Changes in the justification of educational inequalities. The role of perceptions of inequality and meritocracy during the COVID pandemic.

### 2024-12-26 03-methods.qmd
changes methods + analisis

The main data source is the Chilean Longitudinal Social Survey 2016–2022. The ELSOC has been designed as a yearly panel study to evaluate how individuals think, feel, and behave regarding a set of social issues concerning conflict and social cohesion in Chile. The sampling design is probabilistic, stratified, clustered, and multistage. It provides adequate coverage of the country's largest cities (Metropolitan Area of Santiago, Valparaíso, and Concepción) and smaller cities. Consequently, the ELSOC comprises a total of 2,927 participants aged between 18 and 75 years in wave 1. In addition, the sample represents 77% of the total national population and 93% of the urban areas [@elsoc_estudio_2022].

The survey has been conducted yearly since 2016, except for the year 2020, when it was suspended due to the pandemic. The waves 2016, 2017, 2018, 2019 and 2022 were administered using computer-assisted personal interviewing (CAPI). However, a reduced version was conducted using computer-assisted telephone interviews (CATI) in 2021. In addition, wave 3 included a refreshment sample to counter survey attrition. The same sampling strategy as that of wave 1 was implemented for selecting new cases. As a result, the total sample of wave 3 included 3,748 cases, of which 2,229 were part of the original sample and 1,519 were from the refreshment sample. The data from the refreshment sample is not included in the analytical sample because we wanted to analyze longer response trends. Regarding the original sample, the response rate was 62.4% in wave 1, achieving N = 2,927 participants. In broader terms, the accumulated attrition between wave 1 and wave 6 is 41%, achieving a final sample of N = 1728. The analysis used longitudinal weights to avoid bias due to the sampling design, allowing us to control for biases arising from systematic patterns of non-participation in the survey after the first wave. For a more detailed analysis of responses, attrition, and the construction of longitudinal weights, visit https://coes.cl/encuesta-panel/. The dataset is publicly available at: https://dataverse.harvard.edu/dataverse/elsoc.

@@

The dependent variable of this study is the justification of educational inequality. This construct is measured using the following statement: "It is just that high-income people have a better education for their children than people with lower incomes" ("Es justo que las personas de altos ingresos tengan una mejor educación para sus hijos que las personas con ingresos más bajos" in Spanish). Here, the respondents declared their preferences on a Likert scale from "strongly disagree" (1) to “strongly agree” (5). The main independent variables refer to meritocratic perception, which is operationalized by two items, one related to effort ("In Chile, people are rewarded for their efforts"), and the other related to talent ("In Chile, people are rewarded for their intelligence and skills"). In addition, we included an indicator for economic inequality perception ("In Chile, the income differences are too large"). @fig-frecuencies shows the average distribution throughout the years for these variables. We observe that there is a large group (88.2%) of respondents who disagree or strongly disagree that people with higher incomes can access better education. Likewise, a large group of respondents (91.7%) agree or strongly agree that income differences in Chile are too large. On the other hand, there is an important level of disagreement (disagree + strongly disagree) about the perception of meritocracy, this is, whether people are rewarded for their talent (47.6%) and for their efforts (57%).

@@

For testing the hypothesis regarding status we included educational level and household income quintiles, as well as subjective social status. Finally, political identification on a left-right scale, age, and gender were included as control variables. @tbl-descriptives shows the whole set of independent variables, the corresponding items,  response categories, and frequencies.

>>>>

The main data source is the Chilean Longitudinal Social Survey 2016–2023. The ELSOC has been designed as a yearly panel study to evaluate how individuals think, feel, and behave regarding a set of social issues concerning conflict and social cohesion in Chile. The sampling design is probabilistic, stratified, clustered, and multistage. It provides adequate coverage of the country's largest cities (Metropolitan Area of Santiago, Valparaíso, and Concepción) and smaller cities. Consequently, the ELSOC comprises a total of 2,927 participants aged between 18 and 75 years in wave 1. In addition, the sample represents 77% of the total national population and 93% of the urban areas [@elsoc_estudio_2022].

The survey has been conducted yearly since 2016, except for the year 2020, when it was suspended due to the pandemic. The waves 2016, 2017, 2018, 2019 and 2022 were administered using computer-assisted personal interviewing (CAPI). However, a reduced version was conducted using computer-assisted telephone interviews (CATI) in 2021. In addition, wave 3 included a refreshment sample to counter survey attrition. The same sampling strategy as that of wave 1 was implemented for selecting new cases. As a result, the total sample of wave 3 included 3,748 cases, of which 2,229 were part of the original sample and 1,519 were from the refreshment sample. The data from the refreshment sample is not included in the analytical sample because we wanted to analyze longer response trends. Regarding the original sample, the response rate was 62.4% in wave 1, achieving N = 2,927 participants. In broader terms, the accumulated attrition between wave 1 and wave 6 is 41%, achieving a final sample of N = 1737. The analysis used longitudinal weights to avoid bias due to the sampling design, allowing us to control for biases arising from systematic patterns of non-participation in the survey after the first wave. For a more detailed analysis of responses, attrition, and the construction of longitudinal weights, visit https://coes.cl/encuesta-panel/. The dataset is publicly available at: https://dataverse.harvard.edu/dataverse/elsoc.

@@

The dependent variable of this study is the justification of educational inequality. This construct is measured using the following statement: "It is just that high-income people have a better education for their children than people with lower incomes" ("Es justo que las personas de altos ingresos tengan una mejor educación para sus hijos que las personas con ingresos más bajos" in Spanish). Here, the respondents declared their preferences on a Likert scale from (1) "strongly disagree" to (5) “strongly agree” (5). The main independent variables refer to meritocratic perception, which is operationalized by two items, one related to effort ("In Chile, people are rewarded for their efforts"), and the other related to talent ("In Chile, people are rewarded for their intelligence and skills"). In addition, we included an indicator for economic inequality perception ("In Chile, the income differences are too large"). @fig-frecuencies shows the average distribution throughout the years for these variables. We observe that there is a large group (88.2%) of respondents who disagree or strongly disagree that people with higher incomes can access better education. Likewise, a large group of respondents (91.7%) agree or strongly agree that income differences in Chile are too large. On the other hand, there is an important level of disagreement (disagree + strongly disagree) about the perception of meritocracy, this is, whether people are rewarded for their talent (47.6%) and for their efforts (57%).

@@

A third group of variables is included in the analysis to assess the effect of elements related to the pandemic. These variables were only included in the 2021 wave, so each individual's responses were replicated for the remaining years. The first variable addresses the belief regarding whether, in the context of the pandemic, it is more important to give priority to economic activity than to the health of the population, measured on a Likert scale from (1) "strongly disagree" to (2) "strongly agree". The second variable is a dummy (No/Yes) that captures whether individuals received state benefits during the pandemic (e.g., Food Box, Emergency Family Income, Contribution for the Middle Class, or others).

For testing the hypothesis regarding status we included educational level and household income quintiles, as well as subjective social status. Finally, political identification on a left-right scale, age groups, and gender were included as control variables. @tbl-descriptives shows the whole set of independent variables, the corresponding items,  response categories, and frequencies.

### 2024-12-26 04-analisis.qmd
changes methods + analisis

```{r echo=FALSE, fig.align='center'}
#| label: fig-alluvial
#| fig-cap: "Change in the justification of educational inequality over time (2016-2022)"
#| 
 knitr::include_graphics("output/graphs/alluvial_dep.png")
```

@fig-alluvial illustrates yearly frequencies in the justification of educational inequality between 2016 and 2022. Each year represents stacked percentual frequencies, and the flows in between reflect the within-subject change of opinions from one year to the next, as we are using longitudinal panel data [@rosvall_mapping_2010]. For instance, of the 32.2% who strongly disagreed with inequality justification in 2016, about half of them kept responding the same in 2017, whereas the other half shifted their opinion to other response categories. In general, the large majority - between 80 and 90% - disagrees with the justification of educational inequality throughout the years. Despite this overall tendency, we also observe that the disagreement with inequality justification (disagree + strongly disagree) tends to go down in the last wave. This change is mostly a result of the radical decrease in the lowest response category (strongly disagree), which diminishes to half when compared with previous years.

@@

@fig-means shows the average changes in the main variables considered for this study. Here we observe that the justification of educational inequality has the lowest average throughout the years when compared with the other (independent) variables, whereas the highest average is consistently represented by inequality perception. Interestingly, in the last wave of the study (2022) the justification of inequality increases whereas the perception of inequality decreases. As the merit variables are concerned, they show a very similar pattern in terms of averages and changes over the years, being the perception of meritocracy related to effort always lower than the one associated with talent.

@@

@fig-correlation presents a correlation matrix of the main variables analyzed, using data from all survey waves. In this matrix, the correlations vary between low and moderate values. Justification of economic inequality depicts a moderate and negative association with the perception of inequality (_r_=-0.29, p<.01) and a moderate and positive association with both meritocracy perception variables (_r_=0.18, p<.01; _r_=0.15, p<.01). Regarding the perception of inequality, it presents a moderate and negative association with both meritocracy perception variables (_r_=-0.15, p<.01; _r_=-0.10, p<.01). Finally, the two meritocracy variables present a high and positive association with each other (_r_=0.74, p<.01).

@@

@tbl-multilevel shows the multilevel estimation results for the justification of educational inequality. Model 1 includes the survey waves to estimate intertemporal changes in the dependent variable. Taking 2016 as a reference point, we can observe a staggered decrease in 2017 ($\beta$=-0.15, p<.001), 2018 ($\beta$=-0.06, p<.05), and 2019 ($\beta$=-0.07, p<.01). Nevertheless, in the last wave of 2022 there is a radical increase in level of justification of economic inequality ($\beta$=0.21, p<.001), suggesting a non-linear change in this variable. Attempting to model this path of change over time, Model 2 incorporates time (survey waves) as a continuous variable as well as its quadratic term, representing the nonlinear association initially observed in Model 1. On the one hand, the survey wave depicts a negative association, expressing an average decrease in inequality justification over time, but on the other hand, the quadratic wave term is positive, indicating the reversion of this path in the last measurement point.

Model 3 adds the sociodemographic variables, where having a higher income and being older (65 or older compared to 18-29) have a significant and positive influence on the justification of educational inequality. On the contrary, the gender variable has a negative effect, expressing that women on average justify less inequality in education when compared to males. Educational level and subjective social status are not shown in this table for the sake of space (as they do not exhibit significant effects), but the complete estimation of the models is presented in the appendix. Model 3 also includes the control variables, where identifying oneself with the center, right-wing, or when reporting no political identification, has a significant and positive association on the justification of educational inequality - when compared to being left-wing oriented. Inequality perception is added in Model 4, showing a negative association with the justification of educational inequality as hypothesized, remaining stable when controlling for the rest of the variables.

Models 5 and 6 introduce the meritocratic variables: talent (if intelligence and abilities are rewarded in society) and effort (if efforts are rewarded in society). In line with our hypotheses, the perception that talent is rewarded has a positive influence on the justification of educational inequality in Model 5 ($\beta$=0.04, p<.01). However, when controlling for the perception that effort is rewarded, this effect is no longer significant. In this sense, Model 6 shows that the perception that effort is rewarded in society is not only positively associated with the justification of educational inequality ($\beta$=0.05, p<.01), but it has a larger weight than the perception of talent.

@@

# Hipotesis
h0 <- "ola"
h1 <- "ola_num+ola_2"
h2 <- "ola_num+ola_2+education+quintil1+ess+edad+sexo"
h3 <- "ola_num+ola_2+education+quintil1+ess+edad+sexo+pos_id"
h4 <- "ola_num+ola_2+education+quintil1+ess+edad+sexo+pos_id+des_perc"
h5 <- "ola_num+ola_2+education+quintil1+ess+edad+sexo+pos_id+des_perc+merit_talent"
h6 <- "ola_num+ola_2+education+quintil1+ess+edad+sexo+pos_id+des_perc+merit_talent+merit_effort"
h7 <- "ola_num+ola_2+education+quintil1+ess+edad+sexo+pos_id+des_perc+merit_talent+merit_effort+covid_economia"
h8 <- "ola_num+ola_2+education+quintil1+ess+edad+sexo+pos_id+des_perc+merit_talent+merit_effort+covid_economia+covid_beneficios"

@@

omit <- "(education)|(ess)"
texreg::knitreg(list(educ0, educ1, educ3, educ4, educ5, educ6,educ7,educ8),
                custom.note = "*** p < 0.001; ** p < 0.01; * p < 0.05.
Note: Model 3 to 6 are controlled by educational level, and subjective social status with no significant effects.", 
                omit.coef = omit,
                custom.coef.names = c("Intercept",
                                      "Wave 2017",
                                      "Wave 2018",
                                      "Wave 2019",
                                      "Wave 2022",
                                      "Wave 2023",
                                      "Wave",
                                      "Wave^2",
#                                      "High School <br> <i>(Ref. Primary)</i>",
#                                      "Technical",
#                                      "Universitary",
                                      "Quintile 2",
                                      "Quintile 3",
                                      "Quintile 4",
                                      "Quintile 5",
                                      "Quintile (No information)",
#                                      "Subjective Social Status",
                                      "Age 30-49",
                                      "Age 50-64",
                                      "Age 65 or more",
                                      "Gender (Ref. Male)",
                                      "Center",
                                      "Right",
                                      "Does not identify",
                                      "Inequality perception",
                                      "Merit: Talent",
                                      "Merit: Effort" ,
                                      "Covid: economy more important than health",
                                      "Covid: receive state benefits"),
                groups = list("Wave (Ref.= Wave 2016)" = 2:6,
                  "Household Income (Ref.= Quintile 1)" = 9:13,
                  "Age (Ref.= 18-29)" = 14:16,
                  "Pol. pos (Ref.= Left)" = 18:20
                  ),
                include.loglik = FALSE,
                include.aic = FALSE,
                custom.gof.names = c("BIC",
                                     "Num. obs.",
                                     "Num. groups: Individuals",
                                     "Var: Individuals (Intercept)",
                                     "Var: Residual"),
               center = TRUE,
               caption.above = TRUE,
               scalebox = 0.765,
               float.pos = "!ht", 
               booktabs = FALSE, 
               use.packages = FALSE, 
               include.rs = FALSE, 
               include.rmse = FALSE,
               include.ci = FALSE)

@@

```{r interact, echo=FALSE, results='asis'}
#| label: tbl-interact
#| fig-cap: "Interaction effects for the justification of economic inequality"

@@

h6 <- "ola+ola_2+education+quintil1+ess+edad+sexo+pos_id+des_perc+merit_talent+merit_effort+covid_economia+covid_beneficios"
h1b <- "ola+ola_2+education+quintil1+ess+edad+sexo+pos_id+des_perc+merit_talent+merit_effort+covid_economia+covid_beneficios+merit_effort*ola"
h3b <- "ola+ola_2+education+quintil1+ess+edad+sexo+pos_id+des_perc+merit_talent+merit_effort+covid_economia+covid_beneficios+merit_talent*ola"

# A. justification educational inequality
educ1 <- lmer(formula(paste0("just_educ~",h1b,"+(ola+merit_effort|idencuesta)")),data = df_study1, weights = ponderador_long_total)
educ3 <- lmer(formula(paste0("just_educ~",h3b,"+(ola+merit_talent|idencuesta)")),data = df_study1, weights = ponderador_long_total)
educ6 <- lmer(formula(paste0("just_educ~",h6,"+(1|idencuesta)")),data = df_study1, weights = ponderador_long_total)

# table 3: interacciones
omit <- "(education)|(ess)|(quintil1)|(edad)|(sexo)|(pos_id)|(des_perc)|(covid_beneficios)|(covid_economia)"
texreg::knitreg(list(educ6,educ3,educ1),
                custom.note = "\\item *** p < 0.001; ** p < 0.01; * p < 0.05. Note: All the models are controlled by educational level, income quintile, subjective social status, age, gender, political position, and inequality perception",
                omit.coef = omit,
                include.loglik = FALSE,
                include.aic = FALSE,
                custom.coef.names = c("Intercept",
                                      "Wave",
                                      "Wave^2",
                                      "Merit: Talent",
                                      "Merit: Effort",
                                      "Wave * Merit: Talent",
                                      "Wave * Merit: Effort"
                                      ),
                custom.gof.names = c("BIC",
                                     "Num. obs.",
                                     "Num. groups: Individuals",
                                     "Var: Individuals (Intercept)",
                                     "Var: Residual",
                                     "Var: Individuals Wave",
                                     "Var: Individuals Merit: Talent",
                                     "Cov: Individuals (Intercept) Wave",
                                     "Cov: Individuals (Intercept) Merit: Talent",
                                     "Cov: Individuals Wave Merit: Talent",
                                     "Var: Individuals Merit: Effort",
                                     "Cov: Individuals (Intercept) Merit: Effort",
                                     "Cov: Individuals Wave Merit: Effort"),  
               center = TRUE,
               caption.above = TRUE,
               scalebox = 0.765,
               float.pos = "!ht", 
               booktabs = TRUE, 
               use.packages = FALSE, 
               include.rs = FALSE, 
               include.rmse = FALSE,
               include.ci = FALSE,
               threeparttable = TRUE)
```

In this final part of the analysis, we contrast longitudinal hypotheses regarding changes in the relationship between meritocracy and justification of economic inequality over time. In hypothesis 5 we proposed that the association between the perception of meritocracy and inequality justification mitigates over time, as meritocratic ideals could have weakened due to critical situations associated with the COVID pandemic. We test this hypothesis through interaction effects, displayed in @tbl-interact. Model 1 is shown as a baseline model, it is the same as Model 6 in @tbl-interact but it only displays the variables involved in the interaction for the sake of space (all other variables are controlled for). Model 2 adds the interaction between the perception of meritocracy based on talent and time (panel wave), whereas Model 3 does the same but now for the perception of meritocracy related to effort. As observed, only the effort variable shows a significant interaction with the time variable ($\beta$=0.01, p<.05), meaning that, on average, the association between the perception of meritocracy based on effort and the justification of inequality increases by 0.01 points in every measurement point.

>>>>

```{r echo=FALSE, fig.align='center'}
#| label: fig-alluvial
#| fig-cap: "Change in the justification of educational inequality over time (2016-2023)"
#| 
 knitr::include_graphics("output/graphs/alluvial_dep.png")
```

@fig-alluvial illustrates yearly frequencies in the justification of educational inequality between 2016 and 2023. Each year represents stacked percentual frequencies, and the flows in between reflect the within-subject change of opinions from one year to the next, as we are using longitudinal panel data [@rosvall_mapping_2010]. For instance, of the 32.5% who strongly disagreed with inequality justification in 2016, about half of them kept responding the same in 2017, whereas the other half shifted their opinion to other response categories. In general, the large majority - between 80 and 90% - disagrees with the justification of educational inequality throughout the years. Despite this overall tendency, we also observe that the disagreement with inequality justification (disagree + strongly disagree) tends to go down in the last waves. This change is mostly a result of the increase in the "agree" category, which doubles compared to previous years (from 5.4% in 2016 to 11.1% in 2023).

@@

@fig-means shows the average changes in the main variables considered for this study. Here we observe that the justification of educational inequality has the lowest average throughout the years when compared with the other (independent) variables, whereas the highest average is consistently represented by inequality perception. Interestingly, in the last waves of the study (2022-2023) the justification of inequality increases whereas the perception of inequality decreases. As the merit variables are concerned, they show a very similar pattern in terms of averages and changes over the years, being the perception of meritocracy related to effort always lower than the one associated with talent.

@@

@fig-correlation presents a correlation matrix of the main variables analyzed, using data from all survey waves. In this matrix, the correlations vary between low and moderate values. Justification of education inequality depicts a moderate and negative association with the perception of inequality (_r_=-0.29, p<.01) and a moderate and positive association with both meritocracy perception variables (_r_=0.19, p<.01; _r_=0.16, p<.01). Regarding the perception of inequality, it presents a moderate and negative association with both meritocracy perception variables (_r_=-0.15, p<.01; _r_=-0.11, p<.01). Finally, the two meritocracy variables present a high and positive association with each other (_r_=0.74, p<.01).

@@

@tbl-multilevel shows the multilevel estimation results for the justification of educational inequality. Model 1 includes the survey waves to estimate intertemporal changes in the dependent variable. Taking 2016 as a reference point, we can observe a staggered decrease in 2017 ($\beta$=-0.17, p<.001), 2018 ($\beta$=-0.07, p<.05), and 2019 ($\beta$=-0.07, p>.05). Nevertheless, in the last waves of 2022 and 2023 there is a radical increase in level of justification of economic inequality ($\beta$=0.2, p<.001 and $\beta$=0.21, p<.001), suggesting a non-linear change in this variable. Attempting to model this path of change over time, Model 2 incorporates time (survey waves) as a continuous variable as well as its quadratic term, representing the nonlinear association initially observed in Model 1. On the one hand, the survey wave depicts a negative association, expressing an average decrease in inequality justification over time, but on the other hand, the quadratic wave term is positive, indicating the reversion of this path in the last measurements point.

Model 3 adds the socioeconomic variables to test hypothesis 2. Educational level and household income quintiles were added, however, none of them had a significant effect. In addition, from this model onwards the demographic control variables were added, where only being right-wing (compared to being left-wing) had a positive and significant effect ($\beta$=0.27, p<.001). Subjective social status, age and gender had no significant effect when controlling for all the variables analysed.

Models 4 and 5 add the two variables about the Covid context, where we can see that those who believe that in this context the economy is more important than the health of the population justify more inequality in education, an effect that is maintained when controlling for all the variables in model 8 ($\beta$=0.05, p<.01). On the other hand, those who received state benefits during the pandemic justify less inequality in education compared to those who did not receive them, an effect that also remains significant when controlling for the rest of the variables in model 8 ($\beta$=-0.08, p<.01).

Inequality perception is added in Model 6, showing a negative association with the justification of educational inequality as hypothesized, remaining stable when controlling for the rest of the variables. Model 7 and 8 introduce the meritocratic variables: talent (if intelligence and abilities are rewarded in society) and effort (if efforts are rewarded in society). In line with our hypotheses, the perception that talent is rewarded has a positive influence on the justification of educational inequality in Model 7 ($\beta$=0.05, p<.001). However, when controlling for the perception that effort is rewarded, this effect is no longer significant. In this sense, Model 8 shows that the perception that effort is rewarded in society is not only positively associated with the justification of educational inequality ($\beta$=0.08, p<.001), but it has a larger weight than the perception of talent.

@@

# Hipotesis
h0 <- "ola"
h1 <- "ola_num+ola_2"
h2 <- "ola_num+ola_2+education+quintil1+ess+edad+sexo"
h3 <- "ola_num+ola_2+education+quintil1+ess+edad+sexo+pos_id"
h4 <- "ola_num+ola_2+education+quintil1+ess+edad+sexo+pos_id+covid_economia"
h5 <- "ola_num+ola_2+education+quintil1+ess+edad+sexo+pos_id+covid_economia+covid_beneficios"
h6 <- "ola_num+ola_2+education+quintil1+ess+edad+sexo+pos_id+covid_economia+covid_beneficios+des_perc"
h7 <- "ola_num+ola_2+education+quintil1+ess+edad+sexo+pos_id+covid_economia+covid_beneficios+des_perc+merit_talent"
h8 <- "ola_num+ola_2+education+quintil1+ess+edad+sexo+pos_id+covid_economia+covid_beneficios+des_perc+merit_talent+merit_effort"

@@

omit <- "(edad)|(ess)|(sexo)|(pos_id)"
texreg::knitreg(list(educ0, educ1, educ3, educ4, educ5, educ6,educ7,educ8),
                custom.note = "*** p < 0.001; ** p < 0.01; * p < 0.05.
Note: Model 3 to 8 are controlled by gender, political position, and subjective social status with no significant effects.", 
                omit.coef = omit,
                custom.coef.names = c("Intercept",
                                      "Wave 2017",
                                      "Wave 2018",
                                      "Wave 2019",
                                      "Wave 2022",
                                      "Wave 2023",
                                      "Wave",
                                      "Wave^2",
                                      "High School",
                                      "Technical",
                                      "Universitary",
                                      "Quintile 2",
                                      "Quintile 3",
                                      "Quintile 4",
                                      "Quintile 5",
                                      "Quintile (No information)",
#                                      "Subjective Social Status",
                                      # "Age 30-49",
                                      # "Age 50-64",
                                      # "Age 65 or more",
                                      # "Gender (Ref. Male)",
                                      # "Center",
                                      # "Right",
                                      # "Does not identify",
                                      "Covid: economy more important than health",
                                      "Covid: receive state benefits",
                                      "Inequality perception",
                                      "Merit: Talent",
                                      "Merit: Effort"),
                groups = list("Wave (Ref.= Wave 2016)" = 2:6,
                              "Education (Ref.= Primary)" = 9:11,
                  "Household Income (Ref.= Quintile 1)" = 12:16
                  #"Age (Ref.= 18-29)" = 14:16,
                  #"Pol. pos (Ref.= Left)" = 18:20
                  ),
                include.loglik = FALSE,
                include.aic = FALSE,
                custom.gof.rows = list("Controls"= c("No", "No", rep("Yes", 6))),
                custom.gof.names = c("BIC",
                                     "Num. obs.",
                                     "Num. groups: Individuals",
                                     "Var: Individuals (Intercept)",
                                     "Var: Residual"),
               center = TRUE,
               caption.above = TRUE,
               scalebox = 0.765,
               float.pos = "!ht", 
               booktabs = FALSE, 
               use.packages = FALSE, 
               include.rs = FALSE, 
               include.rmse = FALSE,
               include.ci = FALSE)

@@

```{r interact, echo=FALSE, results='asis', warning=FALSE}
#| label: tbl-interact
#| tbl-cap: "Interaction effects for the justification of economic inequality"

@@

h6 <- "ola+ola_2+education+quintil1+ess+edad+sexo+pos_id+covid_economia+covid_beneficios+des_perc+merit_talent+merit_effort"
h1b <- "ola+ola_2+education+quintil1+ess+edad+sexo+pos_id+covid_economia+covid_beneficios+des_perc+merit_talent+merit_effort+merit_effort*ola"
h3b <- "ola+ola_2+education+quintil1+ess+edad+sexo+pos_id+covid_economia+covid_beneficios+des_perc+merit_talent+merit_effort+merit_talent*ola"

# A. justification educational inequality
educ1 <- lmer(formula(paste0("just_educ~",h1b,"+(ola|idencuesta)")),data = df_study1, weights = ponderador_long_total)
educ3 <- lmer(formula(paste0("just_educ~",h3b,"+(ola|idencuesta)")),data = df_study1, weights = ponderador_long_total)
educ6 <- lmer(formula(paste0("just_educ~",h6,"+(1|idencuesta)")),data = df_study1, weights = ponderador_long_total)

# table 3: interacciones
omit <- "(education)|(ess)|(quintil1)|(edad)|(sexo)|(pos_id)|(des_perc)|(covid_beneficios)|(covid_economia)"
texreg::knitreg(list(educ6,educ3,educ1),
                custom.note = "\\item *** p < 0.001; ** p < 0.01; * p < 0.05. Note: All the models are controlled by educational level, income quintile, subjective social status, age, gender, political position, and inequality perception",
                omit.coef = omit,
                include.loglik = FALSE,
                include.aic = FALSE,
                custom.model.names = c("Model 8", "Model 9", "Model 10"),
                custom.coef.names = c("Intercept",
                                      "Wave",
                                      "Wave^2",
                                      "Merit: Talent",
                                      "Merit: Effort",
                                      "Wave * Merit: Talent",
                                      "Wave * Merit: Effort"
                                      ),
                custom.gof.names = c("BIC",
                                     "Num. obs.",
                                     "Num. groups: Individuals",
                                     "Var: Individuals (Intercept)",
                                     "Var: Residual",
                                     "Var: Individuals Wave",
                                     "Cov: Individuals (Intercept) Wave"),  
               center = TRUE,
               caption.above = TRUE,
               scalebox = 0.765,
               float.pos = "!ht", 
               booktabs = TRUE, 
               use.packages = FALSE, 
               include.rs = FALSE, 
               include.rmse = FALSE,
               include.ci = FALSE)
```

In this final part of the analysis, we contrast longitudinal hypotheses regarding changes in the relationship between meritocracy and justification of educational inequality over time. In hypothesis 5 we proposed that the association between the perception of meritocracy and inequality justification mitigates over time, as meritocratic ideals could have weakened due to critical situations associated with the COVID pandemic. We test this hypothesis through interaction effects, displayed in @tbl-interact. Model 1 is shown as a baseline model, it is the same as Model 8 in @tbl-interact but it only displays the variables involved in the interaction for the sake of space (all other variables are controlled for). Model 9 adds the interaction between the perception of meritocracy based on talent and time (panel wave), whereas Model 10 does the same but now for the perception of meritocracy related to effort. As observed, only the talent variable shows a significant interaction with the time variable ($\beta$=-0.01, p<.05), meaning that, on average, the association between the perception of meritocracy based on talent and the justification of inequality decreases by 0.01 points in every measurement point.
